### 2025-07-29 paper/preprint/paper.tex
final pass-through done

Specifically, we propose \textbf{counterfactual training (CT)}: a novel
training regime that leverages counterfactual explanations on-the-fly to
ensure that differentiable models learn plausible and actionable
explanations for the underlying data, while at the same time also being
more robust to adversarial examples (AE). Figure~\ref{fig-poc}
illustrates the outcomes of CT compared to a conventionally trained
model. First, in panel (a), faithful and valid counterfactuals end up
near the decision boundary forming a clearly distinguishable cluster in
the target class (orange). In panel (b), CT is applied to the same
underlying linear classifier architecture resulting in much more
plausible counterfactuals. In panel (c), the classifier is again trained
conventionally and we have introduced a mutability constraint on the
\emph{age} feature at test time---counterfactuals are valid but the
classifier is roughly equally sensitive to both features. By contrast,
the decision boundary in panel (d) has titled, making the model trained
with CT relatively less sensitive to the immutable \emph{age} feature.
To achieve these outcomes, CT draws inspiration from the literature on
contrastive and robust learning: we contrast faithful CEs with
ground-truth data while protecting immutable features, and capitalize on
methodological links between CE and AE by penalizing the model's
adversarial loss on interim (\emph{nascent}) counterfactuals. To the
best of our knowledge, CT represents the first venture in this direction
with promising empirical and theoretical results.

@@

CEs have also been used to improve models in the natural language
processing domain. For example, Wu et al.
(\citeproc{ref-wu2021polyjuice}{2021}) propose \emph{Polyjuice}, a
general-purpose CE generator for language models and demonstrate that
the augmentation of training data with \emph{Polyjuice} improves
robustness in a number of tasks, while Luu and Inoue
(\citeproc{ref-luu2023counterfactual}{2023}) introduce the
\emph{Counterfactual Adversarial Training} (CAT) framework that aims to
improve generalization and robustness of language models by generating
counterfactuals for training samples that are subject to high predictive
uncertainty.

@@

The characterization of faithfulness and plausibility in Def.
\ref{def-explainability} follows Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}), with adapted notation.
Intuitively, plausible counterfactuals are consistent with the data and
faithful counterfactuals are consistent with what the model has learned
about the input data. Ac tionability constraints in Def.
\ref{def-explainability} vary and depend on the context in which
\(\mathbf{M}_\theta\) is deployed. In this work, we choose to only
consider domain and mutability constraints for individual features
\(x_d\) for \(d=1,...,D\). We also limit ourselves to classification
tasks for reasons discussed in Section~\ref{sec-discussion}.

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
generated through gradient descent over \(T\) iterations as originally
proposed by Wachter, Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}). CT adopts
gradient-based CE search in training to generate on-the-fly model
explanations \(\mathbf{x}^\prime\) for the training samples. We use the
term \emph{nascent} to denote interim counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) that have not yet converged. As we
explain below, these nascent counterfactuals can be stored and
repurposed as adversarial examples. Conversely, we consider
counterfactuals \(\mathbf{x}_T^\prime\) as \emph{mature} explanations if
they have either exhausted all \(T\) iterations or converged by reaching
a pre-specified threshold, \(\tau\), for the predicted probability of
the target class:
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.

Formally, we propose the following counterfactual training objective to
train explainable (as in Def. \ref{def-explainability}) models,
\begin{equation}\phantomsection\label{eq-obj}{
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{aligned}
}\end{equation} where \(\text{yloss}(\cdot)\) is any classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing (1) the contrastive divergence, \(\text{div}(\cdot)\),
between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
target class \(y^+\), and (2) the adversarial loss,
\(\text{advloss}(.)\), wrt. nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
of the energy terms involved in \(\text{div}(\cdot)\)
(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). The trade-offs
between these components are adjusted through \(\lambda_{\text{div}}\),
\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\). The full training
regime is sketched out in Algorithm \ref{alg-experiment}.

@@

To generate samples, Grathwohl et al.
(\citeproc{ref-grathwohl2020your}{2020}) use Stochastic Gradient
Langevin Dynamics (SGLD) with an uninformative prior for initialization
but we depart from their methodology: we propose to leverage
counterfactual explainers to generate counterfactuals of observed
training samples. Specifically, we have:
\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
}\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
function defined as
\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\),
with \(y^+\) denoting the index of the randomly drawn target class,
\(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
\(\mathbf{x}_T^\prime\) denotes a mature counterfactual for a randomly
sampled factual from a non-target class generated with a gradient-based
CE generator for up to \(T\) iterations. Intuitively, the gradient of
Equation~\ref{eq-div} decreases the energy of observed training samples
(positive samples) while increasing the energy of counterfactuals
(negative samples) (\citeproc{ref-du2019implicit}{Du and Mordatch
2020}). As the counterfactuals get more plausible (Def.
\ref{def-explainability}) during training, these opposing effects
gradually balance each other out (\citeproc{ref-lippe2024uvadlc}{Lippe
2024}).

@@

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
(\citeproc{ref-balashankar2023improving}{Balashankar et al. 2023};
\citeproc{ref-luu2023counterfactual}{Luu and Inoue 2023}) or adversarial
examples (\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022};
\citeproc{ref-pawelczyk2022exploring}{Pawelczyk et al. 2022}). This
leaves some flexibility with regards to the choice for the
\(\text{advloss}(\cdot)\) term in Equation~\ref{eq-obj}. An intuitive
functional form, but likely not the only sensible choice, is inspired by
adversarial training: \begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with Szegedy et al.
(\citeproc{ref-szegedy2013intriguing}{2014}) who define an adversarial
attack as an ``imperceptible non-random perturbation''. Thus, we work
with a different distinction between CE and AE than Freiesleben
(\citeproc{ref-freiesleben2022intriguing}{2022}) who considers
misclassification as the distinguishing feature of adversarial examples.
One of the key observations of this work is that we can leverage CEs
during training and get AEs essentially for free to reap the
aforementioned benefits of adversarial training.

@@

We make use of nine classification datasets common in the CE/AR
literature. Four of them are synthetic with two classes and different
characteristics: linearly separable clusters (\emph{LS}), overlapping
clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
moons (\emph{Moon}). Next, we have four real-world binary tabular
datasets: \emph{Adult} (Census data) of Becker and Kohavi
(\citeproc{ref-becker1996adult2}{1996}), California housing (\emph{CH})
of Pace and Barry (\citeproc{ref-pace1997sparse}{1997}), Default of
Credit Card Clients (\emph{Cred}) of Yeh
(\citeproc{ref-yeh2016default}{2016}), and Give Me Some Credit
(\emph{GMSC}) from Kaggle (\citeproc{ref-kaggle2011give}{2011}).
Finally, for the convenience of illustration, we use the 10-class
\emph{MNIST} (\citeproc{ref-lecun1998mnist}{LeCun 1998}).

@@

Test accuracies on adversarially perturbed data are shown in
Figure~\ref{fig-acc}. The perturbations size, \(\varepsilon\in[0,0.1]\),
increases along the horizontal axis and includes zero, corresponding to
standard test accuracy for non-perturbed data. For all synthetic
datasets, predictive performance of CT is virtually identical to the
baseline and unaffected by perturbations. For all real-world datasets,
we find that CT substantially improves robustness: while in some cases
baseline accuracies drop to essentially zero for large enough
perturbation sizes, accuracies of CT models remain remarkably robust.

@@

We evaluate the impacts of three types of hyperparameters on CT. In this
section we focus on the highlights and make the full results available
in the supplementary appendix.

@@

Secondly, the strength of the energy regularization,
\(\lambda_{\text{reg}}\) is highly impactful and should be set
sufficiently high to avoid common problems associated with exploding
gradients. The sensitivity with respect to \(\lambda_{\text{div}}\) and
\(\lambda_{\text{adv}}\) is much less evident. While high values of
\(\lambda_{\text{reg}}\) may increase the variability in outcomes when
combined with high values of \(\lambda_{\text{div}}\) or
\(\lambda_{\text{adv}}\), this effect is not particularly pronounced.

@@

\emph{CT increases the training times.} Just like contrastive and robust
learning, CT is more resource-intensive than conventional regimes. Three
factors mitigate this effect: (1) CT yields itself to parallel
execution; (2) it amortizes the cost of CEs for the training samples;
and (3) it can be used to fine-tune conventionally-trained models.

@@

State-of-the-art machine learning models are prone to learning complex
representations that cannot be interpreted by humans. Existing work on
counterfactual explanations has largely focused on designing tools to
generate plausible and actionable for any model. In this work, we
instead hold models accountable for delivering such explanations. We
introduce counterfactual training: a novel training regime that
integrates recent advances in contrastive learning, adversarial
robustness, and CE to incentivize highly-explainable models. Through
theoretical results and extensive experiments, we demonstrate that CT
satisfies this goal while promoting adversarial robustness of models.
Explanations generated from CT-based models are both more plausible
(compliant with the underlying data-generating process) and more
actionable (compliant with user-specified mutability constraints), and
thus meaningful to their recipients. In turn, our work highlights the
value of simultaneously improving models and their explanations.

>>>>

Specifically, we propose \textbf{counterfactual training (CT)}: a novel
training regime that leverages counterfactual explanations on-the-fly to
ensure that differentiable models learn plausible and actionable
explanations for the underlying data, while at the same time also being
more robust to adversarial examples (AE). Figure~\ref{fig-poc}
illustrates the outcomes of CT compared to a conventionally trained
model. First, in panel (a), faithful and valid counterfactuals end up
near the decision boundary forming a clearly distinguishable cluster in
the target class (orange). In panel (b), CT is applied to the same
underlying linear classifier architecture resulting in much more
plausible counterfactuals. In panel (c), the classifier is again trained
conventionally and we have introduced a mutability constraint on the
\emph{age} feature at test time---counterfactuals are valid but the
classifier is roughly equally sensitive to both features. By contrast,
the decision boundary in panel (d) has tilted, making the model trained
with CT relatively less sensitive to the immutable \emph{age} feature.
To achieve these outcomes, CT draws inspiration from the literature on
contrastive and robust learning: we contrast faithful CEs with
ground-truth data while protecting immutable features, and capitalize on
methodological links between CE and AE by penalizing the model's
adversarial loss on interim (\emph{nascent}) counterfactuals. To the
best of our knowledge, CT represents the first venture in this direction
with promising empirical and theoretical results.

@@

CEs have also been used to improve models in the natural language
processing domain. For example, Wu et al.
(\citeproc{ref-wu2021polyjuice}{2021}) propose \emph{Polyjuice}, a
general-purpose CE generator for language models, and demonstrate that
the augmentation of training data with \emph{Polyjuice} improves
robustness in a number of tasks. Luu and Inoue
(\citeproc{ref-luu2023counterfactual}{2023}) introduce the
\emph{Counterfactual Adversarial Training} (CAT) framework that aims to
improve generalization and robustness of language models by generating
counterfactuals for training samples that are subject to high predictive
uncertainty.

@@

The characterization of faithfulness and plausibility in Def.
\ref{def-explainability} follows Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}), with adapted notation.
Intuitively, plausible counterfactuals are consistent with the data and
faithful counterfactuals are consistent with what the model has learned
about the input data. Actionability constraints in Def.
\ref{def-explainability} vary and depend on the context in which
\(\mathbf{M}_\theta\) is deployed. In this work, we choose to only
consider domain and mutability constraints for individual features
\(x_d\) for \(d=1,...,D\). We also limit ourselves to classification
tasks for reasons discussed in Section~\ref{sec-discussion}.

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
generated through gradient descent over \(T\) iterations as originally
proposed by Wachter, Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}). CT adopts
gradient-based CE search in training to generate on-the-fly model
explanations \(\mathbf{x}^\prime\) for the training samples. We use the
term \emph{nascent} to denote interim counterfactuals
\(\mathbf{x}_{\text{AE}}^\prime\) that have not yet converged. As we
explain below, these nascent counterfactuals can be stored and
repurposed as adversarial examples. Conversely, we consider
counterfactuals \(\mathbf{x}_{\text{CE}}^\prime\) as \emph{mature}
explanations if they have converged within the \(T\) iterations by
reaching a pre-specified threshold, \(\tau\), for the predicted
probability of the target class:
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.

Formally, we propose the following counterfactual training objective to
train explainable (as in Def. \ref{def-explainability}) models,
\begin{equation}\phantomsection\label{eq-obj}{
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_{\text{CE}}^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{\text{AE}}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_{\text{CE}}^\prime,y;\theta)
\end{aligned}
}\end{equation} where \(\text{yloss}(\cdot)\) is any classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing (1) the contrastive divergence, \(\text{div}(\cdot)\),
between mature counterfactuals \(\mathbf{x}_{\text{CE}}^\prime\) and
observed samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\)
in the target class \(y^+\), and (2) the adversarial loss,
\(\text{advloss}(.)\), wrt. nascent counterfactuals
\(\mathbf{x}_{\text{AE}}^\prime\). Finally, \(\text{ridge}(\cdot)\)
denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
of the energy terms involved in \(\text{div}(\cdot)\)
(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). The trade-offs
between these components are adjusted through \(\lambda_{\text{div}}\),
\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\). The full training
regime is sketched out in Algorithm \ref{alg-experiment}.

@@

To generate samples, Grathwohl et al.
(\citeproc{ref-grathwohl2020your}{2020}) use Stochastic Gradient
Langevin Dynamics (SGLD) with an uninformative prior for initialization
but we depart from their methodology: we propose to leverage
counterfactual explainers to generate counterfactuals of observed
training samples. Specifically, we have:
\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x}^+,\mathbf{x}_{\text{CE}}^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_{\text{CE}}^\prime,y)
}\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
function defined as
\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\),
with \(y^+\) denoting the index of the randomly drawn target class,
\(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
\(\mathbf{x}_{\text{CE}}^\prime\) denotes a mature counterfactual for a
randomly sampled factual from a non-target class generated with a
gradient-based CE generator for up to \(T\) iterations. Intuitively, the
gradient of Equation~\ref{eq-div} decreases the energy of observed
training samples (positive samples) while increasing the energy of
counterfactuals (negative samples) (\citeproc{ref-du2019implicit}{Du and
Mordatch 2020}). As the counterfactuals get more plausible (Def.
\ref{def-explainability}) during training, these opposing effects
gradually balance each other out (\citeproc{ref-lippe2024uvadlc}{Lippe
2024}).

@@

Based on our analysis in Section~\ref{sec-lit}, counterfactuals
\(\mathbf{x}^\prime\) can be repurposed as additional training samples
(\citeproc{ref-balashankar2023improving}{Balashankar et al. 2023};
\citeproc{ref-luu2023counterfactual}{Luu and Inoue 2023}) or adversarial
examples (\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022};
\citeproc{ref-pawelczyk2022exploring}{Pawelczyk et al. 2022}). This
leaves some flexibility with regards to the choice for the
\(\text{advloss}(\cdot)\) term in Equation~\ref{eq-obj}. An intuitive
functional form, but likely not the only sensible choice, is inspired by
adversarial training: \begin{equation}\phantomsection\label{eq-adv}{
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{\text{AE}}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
}\end{equation} Under this choice, we consider nascent counterfactuals
\(\mathbf{x}_{\text{AE}}^\prime\) as AEs as long as the magnitude of the
perturbation to any single feature is at most \(\varepsilon\). This is
closely aligned with Szegedy et al.
(\citeproc{ref-szegedy2013intriguing}{2014}) who define an adversarial
attack as an ``imperceptible non-random perturbation''. Thus, we work
with a different distinction between CE and AE than Freiesleben
(\citeproc{ref-freiesleben2022intriguing}{2022}) who considers
misclassification as the distinguishing feature of adversarial examples.
One of the key observations of this work is that we can leverage CEs
during training and get AEs essentially for free to reap the
aforementioned benefits of adversarial training.

@@

We make use of nine classification datasets common in the CE/AR
literature. Four of them are synthetic with two classes and different
characteristics: linearly separable Gaussian clusters (\emph{LS}),
overlapping clusters (\emph{OL}), concentric circles (\emph{Circ}), and
interlocking moons (\emph{Moon}). Next, we have four real-world binary
tabular datasets: \emph{Adult} (Census data) of Becker and Kohavi
(\citeproc{ref-becker1996adult2}{1996}), California housing (\emph{CH})
of Pace and Barry (\citeproc{ref-pace1997sparse}{1997}), Default of
Credit Card Clients (\emph{Cred}) of Yeh
(\citeproc{ref-yeh2016default}{2016}), and Give Me Some Credit
(\emph{GMSC}) from Kaggle (\citeproc{ref-kaggle2011give}{2011}).
Finally, for the convenience of illustration, we use the 10-class
\emph{MNIST} (\citeproc{ref-lecun1998mnist}{LeCun 1998}).

@@

Test accuracies on adversarially perturbed data are shown in
Figure~\ref{fig-acc}. The perturbation size, \(\varepsilon\in[0,0.1]\),
increases along the horizontal axis and includes zero, corresponding to
standard test accuracy for non-perturbed data. For all synthetic
datasets, predictive performance of CT is virtually identical to the
baseline and unaffected by perturbations. For all real-world datasets,
we find that CT substantially improves robustness: while in some cases
baseline accuracies drop to essentially zero for large enough
perturbation sizes, accuracies of CT models remain remarkably robust.

@@

We evaluate the impacts of three types of hyperparameters on CT. In this
section, we focus on the highlights and make the full results available
in the supplementary appendix.

@@

Secondly, the strength of the energy regularization,
\(\lambda_{\text{reg}}\), is highly impactful and should be set
sufficiently high to avoid common problems associated with exploding
gradients. The sensitivity with respect to \(\lambda_{\text{div}}\) and
\(\lambda_{\text{adv}}\) is much less evident. While high values of
\(\lambda_{\text{reg}}\) may increase the variability in outcomes when
combined with high values of \(\lambda_{\text{div}}\) or
\(\lambda_{\text{adv}}\), this effect is not particularly pronounced.

@@

\emph{CT increases training times.} Just like contrastive and robust
learning, CT is more resource-intensive than conventional regimes. Three
factors mitigate this effect: (1) CT yields itself to parallel
execution; (2) it amortizes the cost of CEs for the training samples;
and (3) it can be used to fine-tune conventionally-trained models.

@@

State-of-the-art machine learning models are prone to learning complex
representations that cannot be interpreted by humans. Existing work on
counterfactual explanations has largely focused on designing tools to
generate plausible and actionable explanations for any model. In this
work, we instead hold models accountable for delivering such
explanations. We introduce counterfactual training: a novel training
regime that integrates recent advances in contrastive learning,
adversarial robustness, and CE to incentivize highly explainable models.
Through theoretical results and extensive experiments, we demonstrate
that CT satisfies this goal while promoting adversarial robustness of
models. Explanations generated from CT-based models are both more
plausible (compliant with the underlying data-generating process) and
more actionable (compliant with user-specified mutability constraints),
and thus meaningful to their recipients. In turn, our work highlights
the value of simultaneously improving models and their explanations.

### 2025-07-29 paper/sections/intro.qmd
mnist figure untangled
final pass-through done

Specifically, we propose **counterfactual training (CT)**: a novel training regime that leverages counterfactual explanations on-the-fly to ensure that differentiable models learn plausible and actionable explanations for the underlying data, while at the same time also being more robust to adversarial examples (AE). @fig-poc illustrates the outcomes of CT compared to a conventionally trained model. First, in panel (a), faithful and valid counterfactuals end up near the decision boundary forming a clearly distinguishable cluster in the target class (orange). In panel (b), CT is applied to the same underlying linear classifier architecture resulting in much more plausible counterfactuals. In panel (c), the classifier is again trained conventionally and we have introduced a mutability constraint on the *age* feature at test time---counterfactuals are valid but the classifier is roughly equally sensitive to both features. By contrast, the decision boundary in panel (d) has titled, making the model trained with CT relatively less sensitive to the immutable *age* feature. To achieve these outcomes, CT draws inspiration from the literature on contrastive and robust learning: we contrast faithful CEs with ground-truth data while protecting immutable features, and capitalize on methodological links between CE and AE by penalizing the model's adversarial loss on interim (*nascent*) counterfactuals. To the best of our knowledge, CT represents the first venture in this direction with promising empirical and theoretical results.

@@

![Counterfactual explanations (stars) for linear classifiers trained under different regimes on synthetic data: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable. The linear decision boundary is shown in green along with training data colored according to ground-truth labels: $y^-=\text{"loan withheld"}$ (blue) and $y^+=\text{"loan provided"}$ (orange). Class and feature annotations (*debt* and *age*) are for illustrative purposes.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}

>>>>

Specifically, we propose **counterfactual training (CT)**: a novel training regime that leverages counterfactual explanations on-the-fly to ensure that differentiable models learn plausible and actionable explanations for the underlying data, while at the same time also being more robust to adversarial examples (AE). @fig-poc illustrates the outcomes of CT compared to a conventionally trained model. First, in panel (a), faithful and valid counterfactuals end up near the decision boundary forming a clearly distinguishable cluster in the target class (orange). In panel (b), CT is applied to the same underlying linear classifier architecture resulting in much more plausible counterfactuals. In panel (c), the classifier is again trained conventionally and we have introduced a mutability constraint on the *age* feature at test time---counterfactuals are valid but the classifier is roughly equally sensitive to both features. By contrast, the decision boundary in panel (d) has tilted, making the model trained with CT relatively less sensitive to the immutable *age* feature. To achieve these outcomes, CT draws inspiration from the literature on contrastive and robust learning: we contrast faithful CEs with ground-truth data while protecting immutable features, and capitalize on methodological links between CE and AE by penalizing the model's adversarial loss on interim (*nascent*) counterfactuals. To the best of our knowledge, CT represents the first venture in this direction with promising empirical and theoretical results.

@@

![Counterfactual explanations (stars) for linear classifiers trained under different regimes on synthetic data: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable. The linear decision boundary is shown in green along with training data colored according to ground-truth labels: $y^-=\text{"loan withheld"}$ (blue) and $y^+=\text{"loan provided"}$ (orange). Class and feature annotations (*debt* and *age*) are for illustrative purposes.](/paper/figures/poc.svg){#fig-poc fig-env="figure*" width=85%}
